The Concurrent Calculi Formalisation Benchmark

### 2024-02-21 short-paper/short-paper.tex
Minor edits

\institute{
     IT University of Copenhagen, Copenhagen, Denmark \email{maca@itu.dk}, \email{dati@itu.dk}
\and University of Kent, Canterbury, United Kingdom \email{D.Castro-Perez@kent.ac.uk}
\and Royal Holloway, University of London, Egham, United Kingdom \email{Francisco.FerreiraRuiz@rhul.ac.uk}
\and University of Liverpool, Liverpool, United Kingdom \email{Lorenzo.Gheri@liverpool.ac.uk}
\and Technical University of Denmark, Kgs. Lyngby, Denmark \email{fkjac@dtu.dk}, \email{alcsc@dtu.dk}
\and Università degli Studi di Milano, Milan, Italy \email{momigliano@di.unimi.it}
\and Università di Camerino, Camerino, Italy \email{luca.padovani@unicam.it}
\and University of Oxford, Oxford, United Kingdom \email{martin.vassor@cs.ox.ac.uk}, \email{nobuko.yoshida@cs.ox.ac.uk}
\and McGill University, Montreal, Canada \email{daniel.zackon@mcgill.ca}
}

@@

\begin{abstract}
  POPLMark and POPLMark Reloaded sparked a flurry of work on machine-checked proofs, and fostered the adoption of proof mechanisation in programming language research.
  Both challenges were purposely limited in scope, and their benchmark problems do not address concurrency-related issues.
  %
  We propose a new collection of benchmark problems focused on the difficulties that typically arise when mechanising formal models of concurrent and distributed programming languages, such as process calculi.
  Our benchmark problems address three key topics: linearity, scope extrusion, and coinductive reasoning.
  The goal of this new challenge is to clarify, compare, and advance the state of the art, fostering the adoption of proof mechanisation in future research on message-passing concurrency.

@@

In this spirit, we introduce a new collection of benchmark problems
specifically crafted to tackle common challenges encountered during
the mechanisation of formal models for concurrent and distributed
programming languages.
We particularly focus on process calculi.\todo{Why? - they are PL-like but also formal models in themselves}
Concurrent calculi are notably subtle: for instance, it took nine
years before the incorrect proof of type safety in the seminal paper
by Honda et al.~\cite{Honda1998} was rectified in Yoshida \&
Vasconcelos’s subsequent work\cite{Yoshida2007}.
Similarly, several key results in papers on session types have turned out
to be false later on, illustrating the
value of machine-checked proofs~\cite{Gay2020,10.1145/3290343}.

@@

We have begun collecting solutions to our challenges on our website:
%
\begin{center}
  \url{https://concurrentbenchmark.github.io/}
\end{center}
%
In the future, we expect to use the website to promote best practices and tutorials derived from solutions to our challenges.
We encourage everyone to try the challenges using their favourite techniques, and to send us their solutions.

@@

The goal of our challenge on linear reasoning is proving a type safety theorem
for a process calculus with session types, by combining as usual subject
reduction with absence of errors.  We have chosen  only to model linear (as
opposed to shared) channels. Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where a restriction $(\nu ab)\ldots$ binds two dual names $a$
and $b$ as opposite endpoints of a same channel --- and their duality is
reflected in the type system. We model a simple notion of error: a well-typed
processes must never use dual channel endpoints in a non-dual way (\eg by
performing concurrent send/receive operations on a same endpoint, or two
concurrent send operations on both endpoints).
The operational semantics is formulated as a standard reduction relation; hence,
proving subject reduction requires proving type preservation for structural
congruence.

We designed this challenge to focus on linear reasoning while minimising
definitions and other concerns.  Therefore, we forgo channel delegation:
send/receive operations only support values which do not include channels, hence
the topology of the communication network described by a process cannot change.
We do not allow recursion or replication, hence no infinite behaviours can be
expressed. We also forgo more sophisticated notions of error-freedom (\eg
deadlock freedom) because proving them would distract from the core linear
aspects of the challenge.

In mechanised meta-theory, addressing linearity means choosing an appropriate
representation of the linear typing context.  While a linear context is perhaps
best seen as a multiset, most proof assistants have better support for list
processing.  The latter representation is intuitive, but may require
establishing a large number of technical lemmata that are orthogonal to the
problem under study (in our case, proving type safety for
session types).  In this space, several designs are possible: one can label
occurrences of resources to constrain their usage (\eg~\cite{CicconeP20}); some
authors instead impose a multiset structure over lists
(\eg~\cite{Danielsson12,ChaudhuriLR19}). Alternatively, contexts can be
implemented as finite maps (as in \cite{Castro2020}), whose operations are
sensitive to a linear discipline. In all these cases, the effort required to
develop the infrastructure is significant.
%
One alternative strategy is to bypass the problem of context splitting
by adopting familiar ideas from algorithmic linear type checking. This
approach, referred to as ``typing with leftovers'', is exemplified
in~\cite{DBLP:conf/forte/ZalakainD21}.  Whatever the choice,
list-based encodings can be refined to be intrinsically-typed, if the
proof assistant supports dependent types
(see~\cite{Thiemann2019,CicconeP20,RouvoetPKV20}).
%
A radically different approach is to adopt a \emph{substructural}
meta-logical framework, which handles resource
distribution implicitly, including splitting and
substitution: users need only map their linear operations to the
ones offered by the framework.  The only such framework is
\emph{Celf}~\cite{Schack-Nielsen:IJCAR08} (see the encoding of session
types in~\cite{Bock2016}); unfortunately, \emph{Celf} does not yet
fully support the verification of meta-theoretic properties.  A
compromise is the so called \emph{two-level} approach, \ie encoding a
substructural specification logic in a mainstream proof assistant and
then using that logic to state and prove linear properties (for a
recent example, see~\cite{Felty:MSCS21}).

\subsection{Scope extrusion}
This challenge revolves around the mechanisation of % explicit
scope
extrusion, by which a process can send restricted names to another
process, as long as the restriction can safely be extruded to include
the receiving process.  The setting for this challenge is a
``classic'' untyped \( \pi \)-calculus, where (unlike the calculi in
the other challenges) names can be sent and received, and bound by
input constructs.  We define two different semantics for our system:
\begin{enumerate}
\item A reduction system: this avoids explicit reasoning about scope
  extrusion by using instead structural congruence, allowing
  implementors to explore alternative ways to encode the latter (\eg
  via process contexts or via compatible refinement);
\item A (early) labelled transition system.
\end{enumerate}
The objective of this
challenge is to prove that the two semantics are equivalent up to
structural congruence.

This is the challenge most closely related to  POPLMark, 
since it concerns the properties of binders, whose encoding has been
extensively studied w.r.t.\ the $\lambda$-calculi. Process calculi
present additional challenges,  typically including several
different binding constructs: inputs bind a received name or value,
recursive processes bind recursion variables, and restrictions bind
names.  The first two act similarly to the binders known from
$\lambda$-calculi, but restrictions may be more challenging due to
scope extrusion.

Given those peculiarities, the syntax and semantics of $\pi$-calculi have been
mechanised from an early age (see~\cite{Melham1994}) with many proof
assistants and in so many encoding styles that it would take too long to
review.  They range from concrete encodings based on named syntax
\cite{Melham1994} to basic de Bruijn \todo{cite} and locally nameless
representation%~\cite{Chargueraud2012}
(\eg~\cite{Castro2020}).
Nominal approaches are also common, see~\cite{Bengtson2009}, but they
may be problematic w.r.t.~proof assistants based on dependent types.

@@

The goal of our challenge on coinductive reasoning is proving that strong barbed bisimilarity can be turned into a congruence by making it sensitive to substitution and parallel composition.
The setting for our challenge is an untyped \(\pi\)-calculus with replication of processes enabling infinite behaviours.
We do not include delegation in the calculus since it is orthogonal to our primary aim of exploring coinductive proof techniques.
We base our definition of bisimilarity on a labelled transition system semantics and an observability predicate describing the communication steps available to a process.
The introduction of strong barbed bisimulation is one of the first steps when studying the observable behaviour of process calculi, both in classical textbooks (e.g., \cite{picalcbook}) and in existing mechanisations.
Though weak barbed congruence is a more common behavioural equivalence, we prefer strong equivalences to simplify the theory by avoiding the need to abstract over the number of internal transitions in a trace.

@@

% \cite{Brady2017} notes that using algebraic effects is unwieldy when there are many communication channels and suggests an experimental approach using dependent uniqueness types.
% \cite{Zalakain2019} notes that systems lacking linear types must simulate linearity, and that using a linearity predicate complicates congruences.
% They had to limit themselves to an asymmetric structural precongruence to be able to prove subject reduction.
Some authors have already described the difficulties that combining techniques may induce in a mechanisation, e.g.~\cite{DBLP:conf/forte/ZalakainD21}.

Features that are in wide use, but not covered by our benchmark, include choice constructs and recursion.
Some interesting aspects of message-passing calculi include multiparty session types~\cite{10.1145/2827695} and choreographies~\cite{Carbone2013}, as their meta-theory includes aspects -- \eg well-formedness conditions on global types, partiality of end-point projection function, \etc -- that we do not address.
Other areas of interest include higher-order calculi~\cite{Hirsch2022}, conversation types~\cite{DBLP:journals/tcs/CairesV10}, psi-calculi~\cite{lmcs:696}, or encodings between different calculi~\cite{DBLP:journals/iandc/Gorla10,DBLP:journals/iandc/DardhaGS17}.

>>>>

\institute{
     IT University of Copenhagen, Copenhagen, Denmark \email{maca@itu.dk}, \email{dati@itu.dk}
\and University of Kent, Canterbury, United Kingdom \email{d.castro-perez@kent.ac.uk}
\and Royal Holloway, University of London, Egham, United Kingdom \email{francisco.ferreiraruiz@rhul.ac.uk}
\and University of Liverpool, Liverpool, United Kingdom \email{lorenzo.gheri@liverpool.ac.uk}
\and Technical University of Denmark, Kgs. Lyngby, Denmark \email{fkjac@dtu.dk}, \email{alcsc@dtu.dk}
\and Università degli Studi di Milano, Milan, Italy \email{momigliano@di.unimi.it}
\and Università di Camerino, Camerino, Italy \email{luca.padovani@unicam.it}
\and University of Oxford, Oxford, United Kingdom \email{martin.vassor@cs.ox.ac.uk}, \email{nobuko.yoshida@cs.ox.ac.uk}
\and McGill University, Montreal, Canada \email{daniel.zackon@mcgill.ca}
}

@@

\begin{abstract}
  POPLMark and POPLMark Reloaded sparked a flurry of work on machine-checked proofs, and fostered the adoption of proof mechanisation in programming language research.
  Both challenges were purposely limited in scope, and their benchmark problems do not address concurrency-related issues.
  %
  We propose a new collection of benchmark problems focused on the difficulties that typically arise when mechanising formal models of concurrent and distributed programming languages, such as process calculi.
  Our benchmark problems address three key topics: linearity, scope extrusion, and coinductive reasoning.
  The goal of this new challenge is to clarify, compare, and advance the state of the art, fostering the adoption of proof mechanisation in future research on message-passing concurrency. \todo{Broaden `message-passing concurrency' to `concurrency'?}

@@

In this spirit, we introduce a new collection of benchmark problems
specifically crafted to tackle common challenges encountered during
the mechanisation of formal models for concurrent and distributed
programming languages.
We particularly focus on process calculi.\todo{Why? - they are PL-like but also formal models in themselves}
Concurrent calculi are notably subtle: for instance, it took nine
years before the incorrect proof of type safety in the seminal paper
by Honda et al.~\cite{Honda1998} was rectified in Yoshida and
Vasconcelos’s subsequent work\cite{Yoshida2007}.
Similarly, several key results in papers on session types have turned out
to be false later on, illustrating the
value of machine-checked proofs~\cite{Gay2020,10.1145/3290343}.

@@

We have begun collecting solutions to our challenges on our website:
%
\begin{center}
  \url{https://concurrentbenchmark.github.io/}
\end{center}
%
In the future, we expect to use the website to promote best practices and tutorials derived from solutions to our challenges.
We encourage readers to try the challenges using their favourite techniques, and to send us their solutions.

@@

The goal of our challenge on linear reasoning is proving a type safety theorem
for a process calculus with session types, by combining as usual subject
reduction with absence of errors.  We have chosen  only to model linear (as
opposed to shared) channels. Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where a restriction $(\nu ab)$ binds two dual names $a$
and $b$ as opposite endpoints of the same channel; their duality is
reflected in the type system. We model a simple notion of error: a well-typed
processes must never use dual channel endpoints in a non-dual way (\eg by
performing concurrent send/receive operations on the same endpoint, or two
concurrent send operations on both endpoints).
The operational semantics is formulated as a standard reduction relation. Thus,
proving subject reduction requires proving type preservation for structural
congruence.

We designed this challenge to focus on linear reasoning while minimising
definitions and other concerns.  We therefore forgo channel delegation:
send/receive operations only support values that do not include channels, so
the topology of the communication network described by a process cannot change.
We do not allow recursion or replication, so infinite behaviours cannot be
expressed. We also forgo more sophisticated notions of error-freedom (\eg
deadlock freedom) because proving them would distract from the core linear
aspects of the challenge.

In mechanised meta-theory, addressing linearity means choosing an appropriate
representation of the linear typing context.  While a linear context is perhaps
best seen as a multiset, most proof assistants have better support for list
processing.  The latter representation is intuitive, but may require
establishing a large number of technical lemmata that are orthogonal to the
problem under study (in our case, proving type safety for
session types).  In this space, several designs are possible: one can label
occurrences of resources to constrain their usage (\eg~\cite{CicconeP20}); some
authors instead impose a multiset structure over lists
(\eg~\cite{Danielsson12,ChaudhuriLR19}). Alternatively, contexts can be
implemented as finite maps (as in \cite{Castro2020}), whose operations are
sensitive to a linear discipline. In all these cases, the effort required to
develop the infrastructure is significant.
%
One alternative strategy is to bypass the problem of context splitting
by adopting familiar ideas from algorithmic linear type checking. This
approach, referred to as ``typing with leftovers,'' is exemplified
in~\cite{DBLP:conf/forte/ZalakainD21}.  Whatever the choice,
list-based encodings can be refined to be intrinsically-typed, if the
proof assistant supports dependent types
(see~\cite{Thiemann2019,CicconeP20,RouvoetPKV20}).
%
A radically different approach is to adopt a \emph{substructural}
meta-logical framework, which handles resource
distribution implicitly, including splitting and
substitution: users need only map their linear operations to the
ones offered by the framework.  The only such framework is
\emph{Celf}~\cite{Schack-Nielsen:IJCAR08} (see the encoding of session
types in~\cite{Bock2016}); unfortunately, \emph{Celf} does not yet
fully support the verification of meta-theoretic properties.  A
compromise is the so called \emph{two-level} approach, \ie encoding a
substructural specification logic in a mainstream proof assistant and
then using that logic to state and prove linear properties (for a
recent example, see~\cite{Felty:MSCS21}).

\subsection{Scope extrusion}
This challenge revolves around the mechanisation of % explicit
scope
extrusion, by which a process can send restricted names to another
process, as long as the restriction can safely be extruded to include
the receiving process.  The setting for this challenge is a
``classic'' untyped \( \pi \)-calculus, where (unlike the calculi in
the other challenges) names can be sent and received, and bound by
input constructs.  We define two different semantics for our system:
\begin{enumerate}
\item A reduction system: this avoids explicit reasoning about scope
  extrusion by using instead structural congruence, allowing
  implementors to explore alternative ways to encode the latter (\eg
  via process contexts or via compatible refinement);
\item An (early) labelled transition system.
\end{enumerate}
The objective of this
challenge is to prove that the two semantics are equivalent up to
structural congruence.

This is the challenge most closely related to  POPLMark, 
since it concerns the properties of binders, whose encoding has been
extensively studied with respect to the $\lambda$-calculi. Process calculi
present additional challenges,  typically including several
different binding constructs: inputs bind a received name or value,
recursive processes bind recursion variables, and restrictions bind
names.  The first two act similarly to the binders known from
$\lambda$-calculi, but restrictions may be more challenging due to
scope extrusion.

Given those peculiarities, the syntax and semantics of $\pi$-calculi have been
mechanised from an early age (see~\cite{Melham1994}) with many proof
assistants and in so many encoding styles that it would take too long to
review.  They range from concrete encodings based on named syntax
\cite{Melham1994} to basic de Bruijn \todo{cite} and locally nameless
representation %~\cite{Chargueraud2012}
(\eg~\cite{Castro2020}).
Nominal approaches are also common (see~\cite{Bengtson2009}), but they
may be problematic in proof assistants based on dependent types.

@@

The goal of our challenge on coinductive reasoning is proving that strong barbed bisimilarity can be turned into a congruence by making it sensitive to substitution and parallel composition.
The setting for our challenge is an untyped \(\pi\)-calculus with replication of processes enabling infinite behaviours.
We do not include delegation in the calculus since it is orthogonal to our primary aim of exploring coinductive proof techniques.
We base our definition of bisimilarity on a labelled transition system semantics and an observability predicate describing the communication steps available to a process.
The introduction of strong barbed bisimulation is one of the first steps when studying the observable behaviour of process calculi, both in classical textbooks (\eg, \cite{picalcbook}) and in existing mechanisations.
Though weak barbed congruence is a more common behavioural equivalence, we prefer strong equivalences to simplify the theory by avoiding the need to abstract over the number of internal transitions in a trace.

@@

% \cite{Brady2017} notes that using algebraic effects is unwieldy when there are many communication channels and suggests an experimental approach using dependent uniqueness types.
% \cite{Zalakain2019} notes that systems lacking linear types must simulate linearity, and that using a linearity predicate complicates congruences.
% They had to limit themselves to an asymmetric structural precongruence to be able to prove subject reduction.
Some authors have already described the difficulties that combining techniques may induce in a mechanisation, \eg~\cite{DBLP:conf/forte/ZalakainD21}.

Features that are in wide use, but not covered by our benchmark, include choice constructs and recursion.
Some interesting aspects of message-passing calculi include multiparty session types~\cite{10.1145/2827695} and choreographies~\cite{Carbone2013}, as their meta-theory includes aspects---\eg well-formedness conditions on global types, partiality of end-point projection function, \etc---that we do not address.
Other areas of interest include higher-order calculi~\cite{Hirsch2022}, conversation types~\cite{DBLP:journals/tcs/CairesV10}, psi-calculi~\cite{lmcs:696}, or encodings between different calculi~\cite{DBLP:journals/iandc/Gorla10,DBLP:journals/iandc/DardhaGS17}.
